---
title: "Portfolio Optimisation with a Genetic Algorithm"
author: "leon5354"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: yes
    keep_tex: no
    toc: yes
    toc_depth: 2
urlcolor: blue
---

```{r setup, include=FALSE}
# Render from project root so paths like "data/cache/..." resolve whether the
# report is opened in RStudio (cwd = report/) or via rmarkdown::render().
project_root <- rprojroot::find_root(rprojroot::has_file("README.md"))
knitr::opts_knit$set(root.dir = project_root)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4, fig.align = "center")
options("getSymbols.warning4.0" = FALSE)
```

## Introduction

This report explores portfolio weight optimisation using a genetic algorithm
(GA). Given a universe of stocks, the goal is to decide what fraction of the
portfolio to allocate to each stock so that an objective (Sharpe ratio, raw
return, or risk) is optimised, subject to the constraint that the weights sum
to one and each individual weight lies in $[0, 1]$.

The analysis is split into two parts:

* **Part 1** — start with a hand-picked 10-stock basket and compare four
  fitness functions (max-Sharpe, max-return, min-risk, and two scaled Sharpe
  variants).
* **Part 2** — start from a 50-stock universe and use a binary GA to pick the
  10 best stocks by Sharpe ratio, then re-optimise weights.

Portfolios trained on the 2020–2021 period are then re-evaluated on the
out-of-sample 2021–2022 period as a simple sanity check.

\bigskip
\noindent\textbf{Disclaimer:} This is a learning exercise. Nothing here is
investment advice, and the "validation" approach is far too simplistic to
support any real trading decision.

## Data

Daily returns are loaded from a snapshot cached in `data/cache/` (see
`data/fetch_data.R` for how to refresh it). As of 2024, the
`quantmod::getSymbols` Yahoo Finance path no longer works, so the snapshots
are committed to keep the report runnable.

```{r load-data, echo=TRUE}
suppressMessages({ library(GA); library(xts); library(ggplot2); library(scales);
                   library(forcats); library(tidyr); library(dplyr) })

cache <- "data/cache"
myRetData        <- readRDS(file.path(cache, "returns_10_2020_2021.rds"))
myRetData        <- myRetData[-1, ]   # drop the first all-NA row from diff()
myRetData_future <- readRDS(file.path(cache, "returns_10_2021_2022.rds"))[-1, ]
myRetData_fif    <- readRDS(file.path(cache, "returns_50_2020_2021.rds"))[-1, ]
```

The Part-1 basket covers a mix of ETFs, large-cap tech, retail, and
crypto-equity exposure:

```{r basket-list, echo=FALSE}
cat(paste(colnames(myRetData), collapse = ", "))
```

A few tickers from the original project (notably `JWN`, `DISCA`, `SQ`) are no
longer available through the Yahoo free API and are dropped automatically.

## Part 1: optimising weights on a fixed basket

### Constraint and objective

Because the `GA` package performs maximisation on a single objective, we
collapse return and risk into a single fitness. For the maximum-Sharpe variant
(with a 0% risk-free rate):

```{r sharpe-fn, echo=TRUE}
sharpe <- function(x) {
  mean(x * myRetData) / sqrt(t(x) %*% cov(myRetData) %*% x)
}
```

The full-investment and box constraints are enforced through a quadratic
penalty rather than hard bounds, so the GA can search freely in $[0, 1]^n$:

```{r weight-penalty, echo=TRUE}
all_weight <- function(x) {
  w <- (sum(x) - 1)^2
  for (i in seq_along(x)) {
    w <- w + max(c(0, x[i] - 1))^2 + max(c(0, -x[i]))^2
  }
  w
}
```

The fitness combines the two. `-sharpe` because GA maximises:

```{r obj-fn, echo=TRUE}
obj <- function(x) -sharpe(x) + 100 * all_weight(x)
```

### Maximum-Sharpe portfolio

```{r ga-sharpe, include=FALSE}
ga_res <- ga(
  type = "real-valued", fitness = function(x) -obj(x),
  lower = rep(0, ncol(myRetData)), upper = rep(1, ncol(myRetData)),
  maxiter = 500, run = 100, monitor = FALSE, seed = 1
)
sol <- as.vector(summary(ga_res)$solution)
names(sol) <- colnames(myRetData)
```

The GA converges with fitness `r round(summary(ga_res)$fitness, 5)`, giving the
following weight allocation:

```{r sharpe-solution, echo=FALSE}
sol_df <- data.frame(stock = names(sol), weight = sol)
sol_df <- sol_df[order(-sol_df$weight), ]
knitr::kable(sol_df, digits = 4,
             col.names = c("Stock", "Weight"),
             row.names = FALSE,
             caption = "Maximum-Sharpe portfolio weights")
```

### Comparing fitness functions

To understand how the choice of objective shapes the portfolio, we run four
GA variants:

1. **Max return** — ignore risk, maximise mean return.
2. **Min risk** — minimise portfolio volatility.
3. **More return** — Sharpe ratio with risk scaled down by 0.5 (rewards return).
4. **Less risky** — Sharpe ratio with risk scaled up by 1.5 (penalises risk).

```{r alt-objectives, include=FALSE}
return_max <- function(x) mean(x * myRetData)
obj_max    <- function(x) -return_max(x) + 100 * all_weight(x)

return_least_risk <- function(x) sqrt(t(x) %*% cov(myRetData) %*% x)
obj_least_risk    <- function(x) return_least_risk(x) + 100 * all_weight(x)

sharpe_more_return <- function(x) {
  risk <- sqrt(t(x) %*% cov(myRetData) %*% x)
  mean(x * myRetData) / (risk * 0.5)
}
obj_more_return <- function(x) -sharpe_more_return(x) + 100 * all_weight(x)

sharpe_less_risky <- function(x) {
  risk <- sqrt(t(x) %*% cov(myRetData) %*% x)
  mean(x * myRetData) / (risk * 1.5)
}
obj_less_risky <- function(x) -sharpe_less_risky(x) + 100 * all_weight(x)

ga_max    <- ga(type = "real-valued", fitness = function(x) -obj_max(x),
                lower = rep(0, ncol(myRetData)), upper = rep(1, ncol(myRetData)),
                maxiter = 500, run = 100, monitor = FALSE, seed = 1)
ga_least  <- ga(type = "real-valued", fitness = function(x) -obj_least_risk(x),
                lower = rep(0, ncol(myRetData)), upper = rep(1, ncol(myRetData)),
                maxiter = 500, run = 100, monitor = FALSE, seed = 1)
ga_more   <- ga(type = "real-valued", fitness = function(x) -obj_more_return(x),
                lower = rep(0, ncol(myRetData)), upper = rep(1, ncol(myRetData)),
                maxiter = 500, run = 100, monitor = FALSE, seed = 1)
ga_less   <- ga(type = "real-valued", fitness = function(x) -obj_less_risky(x),
                lower = rep(0, ncol(myRetData)), upper = rep(1, ncol(myRetData)),
                maxiter = 500, run = 100, monitor = FALSE, seed = 1)
```

```{r alt-solution-table, echo=FALSE}
all_solutions <- data.frame(
  stock       = colnames(myRetData),
  max_sharpe  = as.vector(summary(ga_res)$solution),
  max_return  = as.vector(summary(ga_max)$solution),
  min_risk    = as.vector(summary(ga_least)$solution),
  more_return = as.vector(summary(ga_more)$solution),
  less_risky  = as.vector(summary(ga_less)$solution)
)
knitr::kable(all_solutions, digits = 3,
             caption = "Portfolio weights under each objective")
```

```{r plot-solutions, echo=FALSE, fig.height=5, fig.cap="Optimal weights under each objective variant."}
plot_df <- tidyr::pivot_longer(all_solutions, -stock, names_to = "objective",
                               values_to = "weight")
plot_df$objective <- factor(plot_df$objective,
  levels = c("max_sharpe", "max_return", "more_return", "less_risky", "min_risk"),
  labels = c("Max Sharpe", "Max return", "More return", "Less risky", "Min risk"))
ggplot(plot_df, aes(x = stock, y = weight, fill = objective)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(labels = percent) +
  labs(x = NULL, y = "Weight") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

As expected, the **max-return** portfolio concentrates on the highest-mean
stocks, the **min-risk** portfolio spreads weight more evenly, and the scaled
Sharpe variants sit in between.

### Out-of-sample sanity check

We compare each portfolio's Sharpe ratio on the 2020–2021 training period
against its Sharpe ratio on the held-out 2021–2022 period.

```{r perf-functions, echo=TRUE}
allratio <- function(x, data) {
  port_returns <- mean(x * data)
  port_risk    <- sqrt(t(x) %*% cov(data) %*% x)
  c(sharpe = port_returns / port_risk,
    risk   = port_risk,
    return = port_returns)
}

weight_equal <- rep(1 / ncol(myRetData), ncol(myRetData))
```

```{r perf-table, echo=FALSE}
results <- data.frame(
  portfolio = c("Max Sharpe", "Max return", "Min risk",
                "More return", "Less risky", "Equal weight"),
  rbind(
    allratio(as.numeric(sol),                         myRetData),
    allratio(as.vector(summary(ga_max)$solution),     myRetData),
    allratio(as.vector(summary(ga_least)$solution),   myRetData),
    allratio(as.vector(summary(ga_more)$solution),    myRetData),
    allratio(as.vector(summary(ga_less)$solution),    myRetData),
    allratio(weight_equal,                            myRetData)
  ),
  rbind(
    allratio(as.numeric(sol),                         myRetData_future),
    allratio(as.vector(summary(ga_max)$solution),     myRetData_future),
    allratio(as.vector(summary(ga_least)$solution),   myRetData_future),
    allratio(as.vector(summary(ga_more)$solution),    myRetData_future),
    allratio(as.vector(summary(ga_less)$solution),    myRetData_future),
    allratio(weight_equal,                            myRetData_future)
  )
)
colnames(results) <- c("portfolio",
                       paste0(c("sharpe", "risk", "return"), "_train"),
                       paste0(c("sharpe", "risk", "return"), "_test"))
knitr::kable(results, digits = 3,
             caption = "In-sample (2020-2021) vs out-of-sample (2021-2022) performance")
```

Unsurprisingly, the best in-sample Sharpe does **not** translate into the best
out-of-sample Sharpe — 2020 had an unusual combination of strong momentum and
low correlation that simply did not persist. The min-risk portfolio degrades
the least, consistent with the usual finding that volatility is more
persistent than mean return.

## Part 2: selecting stocks with a binary GA

Instead of hand-picking 10 stocks, we use a binary GA to pick 10 out of a
50-stock universe. Each bit indicates whether a stock is included; the fitness
is the sum of Sharpe ratios of the selected stocks, subject to a cardinality
constraint of exactly 10.

```{r universe-sharpe, include=FALSE}
mean_return_by_stocks <- colSums(myRetData_fif, na.rm = TRUE) / 253
weight_fif            <- rep(1, ncol(myRetData_fif))
fif_risk              <- sqrt(t(weight_fif) %*% cov(myRetData_fif) %*% weight_fif)
sharpe_ratio_of_fif   <- mean_return_by_stocks / as.numeric(fif_risk)

stocks_backpack <- data.frame(
  item   = colnames(myRetData_fif),
  sr     = as.numeric(sharpe_ratio_of_fif),
  weight = 1
)
stocks_limit <- 10
evalFunc <- function(x) {
  best_ten_sharpe <- x %*% stocks_backpack$sr
  current_amount  <- x %*% stocks_backpack$weight
  if (current_amount > stocks_limit) 0 else best_ten_sharpe
}

set.seed(1)
GA_pick <- ga("binary", fitness = evalFunc, nBits = nrow(stocks_backpack),
              maxiter = 500, popSize = 30, run = 100, monitor = FALSE)
# GA_pick@solution is a matrix whose rows are equally-good solutions; use the
# first row to extract the chosen stocks.
picked_mask <- as.logical(GA_pick@solution[1, ])
top_ten     <- stocks_backpack[picked_mask, ]
```

The binary GA selects the following 10 stocks from the 50-name universe
(ranked by their individual Sharpe ratios):

```{r top10-table, echo=FALSE}
top_ten_sorted <- top_ten[order(-top_ten$sr), c("item", "sr")]
knitr::kable(top_ten_sorted, digits = 3,
             col.names = c("Selected stock", "Individual Sharpe"),
             row.names = FALSE,
             caption = "Top-10 stocks chosen by the binary GA")
```

We then re-optimise the weights of those 10 stocks using the same max-Sharpe
GA from Part 1.

```{r top10-weights, include=FALSE}
myStocks_top_ten  <- top_ten_sorted$item
myRetData_top_ten <- myRetData_fif[, myStocks_top_ten, drop = FALSE]

# Some columns may have NAs from gaps; restrict to rows with complete data.
good_rows          <- apply(myRetData_top_ten, 1, function(r) all(!is.na(r)))
myRetData_top_ten  <- myRetData_top_ten[good_rows, ]

sharpe_ten <- function(x) {
  mean(x * myRetData_top_ten) / sqrt(t(x) %*% cov(myRetData_top_ten) %*% x)
}
all_weight_ten <- function(x) {
  w <- (sum(x) - 1)^2
  for (i in seq_along(x)) {
    w <- w + max(c(0, x[i] - 1))^2 + max(c(0, -x[i]))^2
  }
  w
}
obj_ten <- function(x) -sharpe_ten(x) + 100 * all_weight_ten(x)

ga_res_top_ten <- ga(
  type = "real-valued", fitness = function(x) -obj_ten(x),
  lower = rep(0, ncol(myRetData_top_ten)),
  upper = rep(1, ncol(myRetData_top_ten)),
  maxiter = 500, run = 100, monitor = FALSE, seed = 1
)
sol_ten <- as.vector(summary(ga_res_top_ten)$solution)
names(sol_ten) <- colnames(myRetData_top_ten)
```

```{r top10-weights-table, echo=FALSE}
sol_ten_df <- data.frame(stock = names(sol_ten), weight = sol_ten)
sol_ten_df <- sol_ten_df[order(-sol_ten_df$weight), ]
knitr::kable(sol_ten_df, digits = 4,
             col.names = c("Stock", "Weight"),
             row.names = FALSE,
             caption = "Optimised weights for the GA-selected top 10")
```

Because the 10 were chosen specifically for their historical Sharpe ratios, the
in-sample Sharpe of this portfolio is higher than the hand-picked Part-1
portfolio. Out-of-sample, however, it suffers from the same 2020-specific
regime change — possibly more so, since picking "winners" from one period is a
classic source of overfitting.

## Discussion

A few takeaways from the exercise:

* The genetic algorithm works fine for portfolio weight optimisation given a
  sensible penalty-based constraint encoding. It is not the most efficient
  tool — convex optimisation (e.g. the `quadprog` package) would solve the
  min-variance and max-Sharpe problems exactly and quickly — but it is
  flexible and makes the trade-off between objectives easy to explore.
* The choice of objective dominates everything else. Max-return, min-risk, and
  Sharpe variants produce visibly different portfolios from the same universe.
* Out-of-sample, none of the in-sample optima survive intact. The min-risk
  portfolio degrades the least, which is consistent with the broader empirical
  finding that covariance is more persistent than mean returns.

## Reproducibility

Refresh the data with `Rscript data/fetch_data.R`. Re-render the report with
`rmarkdown::render("report/portfolio-ga-analysis.Rmd")`.

```{r session-info, echo=FALSE}
si <- sessionInfo()
loaded <- c("GA", "xts", "ggplot2", "quantmod", "tidyverse")
loaded <- loaded[loaded %in% c(names(si$loadedOnly), names(si$otherPkgs))]
cat("R version:", si$platform, "\n")
cat("Packages loaded:", paste(loaded, collapse = ", "), "\n")
```
